---
output:
  html_document: default
  pdf_document: default
---
# Introduction {#intro}

You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```


## About the Data:

**data/04_remove_underscores_average_replicates/04_SC-expression.csv**
1st col "gene": gene names, lnc= long non-coding RNA
all others: experimental IDs. 87 total. See /03_remove_zero_genes/03_condition_annotation.csv
values: expression values. Normalized to TPM 

**data/03_remove_zero_genes/03_condition_annotation.csv**
1st "col: experimental ID. Grouped together by the first 2 letters.
2nd col "primary": strain used, includes WT, temp sensitivity? (ex. 15 deg, 37 deg), test-control, strain codes
3rd col "secondary": experimental test, ex. salinty, temperature, ethanol, untreated...


**data/03_remove_zero_genes/tiddy_expressions.csv**


**data/tidy_data/tiddy_attributes.csv**

**data/tidy_data/tiddy_conditionss.csv**

**data/tidy_data/tiddy_expressions.csv**



